---
title: "Case Study 2 AKSTA Statistical Computing"
author:
  - Hanna Kienast
  - Iulia Mihaela Enache
  - Kateryna Ponomarenko
date: "2024-04-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Maybe include dataset description here.

## Task

# a
Load in R the following data sets which you can find in TUWEL. For each data set, ensure that missing
values are read in properly, that column names are unambiguous. Each data set should contain at the end
only two columns: country and the variable.

rawdata_343.txt which contains the (estimated) median age per country. Pay attention! The delimiter
 is 2 or more white spaces (one space would not work as it would separate country names which contain
 a space); you have to skip the first two lines. Hint: you can look into function read.fwf or the readr
 corresponding function. It might also be useful to use tidyr functions to unite some columns back or
 separate them.
```{r, echo=TRUE}
library(readr)
library(tidyr)
library(dplyr)
library(readxl)
library(data.table)
require(stringr)
```  
 
```{r, echo=TRUE}
median_age_data <- read.fwf("rawdata_343.txt", widths = c(8,66,4,18), skip=2) 
names(median_age_data) <- c("Rank", "Country", "Median_Age")
median_age_data <- median_age_data[, c("Country", "Median_Age")]
median_age_data
```

```{r, echo=TRUE}
# fwf_widths <- fwf_empty("rawdata_347.txt", skip = 2)
# fwf_widths did not work
migration_data <- read.fwf("rawdata_347.txt", widths = c(8,64,6,18), skip=2)
names(migration_data) <- c("Rank", "Country", "Migration_rate", "Date")
migration_data <- migration_data[, c("Country", "Migration_rate")]
migration_data
```

```{r, echo=TRUE}
unemployment_data <- read.csv("rawdata_373.csv")
names(unemployment_data) <- c("Country", "Youth_unemployment_rate")
unemployment_data
```
# b
Merge the data sets containing raw data using dplyr function on the unique keys. Keep the union of all
observations in the tables. What key are you using for merging? Return the dimension of the merged data
set.
I use the column "country" for merging - I encountered an issue where the column for migration rate was always NaN so I had to trim the whitespaces.
 
```{r, echo=TRUE}
median_age_data$Country <- trimws(median_age_data$Country)
migration_data$Country <- trimws(migration_data$Country)
unemployment_data$Country <- trimws(unemployment_data$Country)

joined <- left_join(median_age_data, unemployment_data, by = join_by(Country))
joined_data <- left_join(joined, migration_data, by = join_by(Country == Country))
head(joined_data)
```
# c

You will acquire more country level information such as the classification of the country based on income.
Such an information can be found at https://datahelpdesk.worldbank.org/knowledgebase/articles/906519.
From there extract the classification for 2020 into low/lower-middle/upper-middle/high income countries
 
```{r, echo=TRUE}
income_data <- read_excel("OGHIST.xlsx", sheet="Country Analytical History", col_names = TRUE)
income_data <- income_data[, c("World Bank Analytical Classifications", "...36")]
income_data <- income_data[11:nrow(income_data), ] 
names(income_data) <- c("Country", "Income_class")
income_data
```
# d
Merge this information to the data set in b.
 1. What are the common variables? Can you merge using them? Why or why not?
 I can merge the dataframes on the column Country.
```{r, echo=TRUE}
data <- left_join(joined_data, income_data, by = join_by(Country == Country))
data
```
 
 2. A reliable merging for countries are ISO codes as they are standardized across data sources. Download the
 mapping of ISO codes to countries from https://www.cia.gov/the-world-factbook/references/country
data-codes/ and load it into R.

```{r, echo=TRUE}
#using read.csv
#country_code <- read.csv("Country Data Codes.csv", sep = ",", quote = "", stringsAsFactors = FALSE, header = TRUE, fill=TRUE)
#country_code

#using read_csv
#country_code <- read_csv("Country Data Codes.csv", col_types = , show_col_types = FALSE,   trim_ws = TRUE)
#country_code

#using fread
country_code <- fread("Country Data Codes.csv", sep = ",", header = TRUE, fill=TRUE, stringsAsFactors = FALSE, quote = "")
names(country_code)[names(country_code) == 'ISO 3166'] <- 'ISO'
names(country_code)[names(country_code) == 'Name'] <- 'Country'

#new idea - merge columns when col2 contains country name data
country_code <- country_code %>%
  mutate(
    Country = ifelse(
      str_detect(GENC, "[a-z]"),  # check for lowercase letters
      paste(Country, GENC, sep = ", "),  # Concatenate Name and GENC if condition is true
      Country  # Keep NAme
    ),
    GENC = ifelse(
      str_detect(GENC, "[a-z]"),
      NA,  # Clear GENC if merged
      GENC
    )
  )

#happy bc we only need Name and ISO columns so i dont have to shift the others

country_code <- country_code[, c("Country", "ISO")]

country_code <- country_code %>%
  mutate(across(everything(), ~ str_replace_all(., '"', '')))

country_code
```

 3. Merge the data sets using the ISO codes.
```{r, echo=TRUE}
country_code$Country <- trimws(country_code$Country)
country_code$ISO <- trimws(country_code$ISO)

data <- left_join(data, country_code, by = join_by(Country == Country))
names(data)[names(data) == 'ISO.y'] <- 'ISO'
data$ISO.x <- NULL
data
```
# e
Introduce into the data set information on continent for each country and subcontinent (region). You should
find a way to gather this data. You can find an appropriate online resource, download the data and merge
the information with the existing data set. Name the merged data set df_vars.
```{r, echo=TRUE}
# I found a dataset on kaggle: https://www.kaggle.com/datasets/andradaolteanu/country-mapping-iso-continent-region?resource=download
continents <- read.csv("continents2.csv")
continents
```

```{r, echo=TRUE}

```

