---
title: "Case Study 2 AKSTA Statistical Computing"
author:
  - Hanna Kienast
  - Iulia Mihaela Enache
  - Kateryna Ponomarenko
date: "2024-04-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Task

# a
Load in R the following data sets which you can find in TUWEL. For each data set, ensure that missing
values are read in properly, that column names are unambiguous. Each data set should contain at the end
only two columns: country and the variable.
 
```{r, echo=TRUE}
library(readr)
library(tidyr)
library(dplyr)
library(readxl)
library(data.table)
require(stringr)
```  
 
```{r, echo=TRUE}
median_age_data <- read.fwf("rawdata_343.txt", widths = c(8,66,4,18), skip=2) 
names(median_age_data) <- c("Rank", "Country", "Median_Age")
median_age_data <- median_age_data[, c("Country", "Median_Age")]
head(median_age_data)
```

```{r, echo=TRUE}

migration_data <- read.fwf("rawdata_347.txt", widths = c(8,64,6,18), skip=2)
names(migration_data) <- c("Rank", "Country", "Migration_rate", "Date")
migration_data <- migration_data[, c("Country", "Migration_rate")]
head(migration_data)
```

```{r, echo=TRUE}
unemployment_data <- read.csv("rawdata_373.csv")
names(unemployment_data) <- c("Country", "Youth_unemployment_rate")
head(unemployment_data)
```
# b
Merge the data sets containing raw data using dplyr function on the unique keys. Keep the union of all
observations in the tables. What key are you using for merging? Return the dimension of the merged data
set.
We use the column "country" for merging. We encountered an issue where the column for migration rate was always NaN so we had to trim the whitespaces.
 
```{r, echo=TRUE}
median_age_data$Country <- trimws(median_age_data$Country)
migration_data$Country <- trimws(migration_data$Country)
unemployment_data$Country <- trimws(unemployment_data$Country)

joined <- left_join(median_age_data, unemployment_data, by = join_by(Country))
joined_data <- left_join(joined, migration_data, by = join_by(Country == Country))
head(joined_data)
dim(joined_data)
```
# c
You will acquire more country level information such as the classification of the country based on income.
Such an information can be found at https://datahelpdesk.worldbank.org/knowledgebase/articles/906519.
From there extract the classification for 2020 into low/lower-middle/upper-middle/high income countries
 
```{r, echo=TRUE}
income_data <- read_excel("OGHIST.xlsx", sheet="Country Analytical History", col_names = TRUE)
income_data <- income_data[, c("World Bank Analytical Classifications", "...36")]
income_data <- income_data[11:nrow(income_data), ] 
names(income_data) <- c("Country", "Income_class")
head(income_data)
```
# d
Merge this information to the data set in b.
 1. What are the common variables? Can you merge using them? Why or why not?
 
 We can merge the dataframes on the column Country.
```{r, echo=TRUE}
data <- left_join(joined_data, income_data, by = join_by(Country == Country))
head(data)
```
 
 2. A reliable merging for countries are ISO codes as they are standardized across data sources. Download the
 mapping of ISO codes to countries from https://www.cia.gov/the-world-factbook/references/country
data-codes/ and load it into R.

```{r, echo=TRUE}
country_code <- fread("Country Data Codes.csv", sep = ",", header = TRUE, fill=TRUE, stringsAsFactors = FALSE, quote = "")
names(country_code)[names(country_code) == 'ISO 3166'] <- 'ISO'
names(country_code)[names(country_code) == 'Name'] <- 'Country'

country_code <- country_code %>%
  mutate(
    Country = ifelse(
      str_detect(GENC, "[a-z]"),  
      paste(Country, GENC, sep = ", "),
      Country  
    ),
    GENC = ifelse(
      str_detect(GENC, "[a-z]"),
      NA,  
      GENC
    )
  )

country_code <- country_code[, c("Country", "ISO")]

country_code <- country_code %>%
  mutate(across(everything(), ~ str_replace_all(., '"', '')))

head(country_code)
```

 3. Merge the data sets using the ISO codes.
```{r, echo=TRUE}
country_code$Country <- trimws(country_code$Country)
country_code$ISO <- trimws(country_code$ISO)

data <- left_join(data, country_code, by = join_by(Country == Country))
names(data)[names(data) == 'ISO.y'] <- 'ISO'
data$ISO.x <- NULL
head(data)
```
# e
Introduce into the data set information on continent for each country and subcontinent (region). You should
find a way to gather this data. You can find an appropriate online resource, download the data and merge
the information with the existing data set. Name the merged data set df_vars.

We found this dataset on kaggle: https://www.kaggle.com/datasets/andradaolteanu/country-mapping-iso-continent-region?resource=download
However, despite ISO being standardized, we could not find a dataset where the ISO layout was the same as the one provided from the previous link.
For making the merging work, we had to merge 3 columns to create a similar column to ISO from the previous dataframe.
```{r, echo=TRUE}

continents <- read.csv("continents2.csv")

continents <- continents %>%
  mutate(
    country.code = str_pad(country.code, width = 3, pad = "0"),
        
    ISO = paste(alpha.2, alpha.3, country.code, sep = "|")
    )
  
names(continents)[names(continents) == 'name'] <- 'Country'
continents <- continents[, c("Country", "region", "sub.region", "ISO")]
head(continents)
```

```{r, echo=TRUE}

df_vars <- left_join(data, continents, by = join_by(Country == Country))
names(df_vars)[names(df_vars) == 'ISO.x'] <- 'ISO'
df_vars$ISO.y <- NULL
dim(df_vars) 
head(df_vars)

```

# f
Discuss on the tidyness of the data set df_vars. What are the observational units, what are the variables? What can be considered fixed vs measured variables? Tidy the data if needed.

The obtained data set df_vars is almost tidy, while it almost fully satisfies the requirements to the data to be considered tidy: our data set is the collection of quantitative and qualitative values and they are organized in a way, that each value belongs to an observation and a variable. Our data is organized in a way that each row represents an observational unit (a country) and each column is a variable. All our variables have the same unit and measure the same attribute.

In our case the fixed variables are Country (char), region (char), sub.region (char), ISO (char), Income_class (char). The left variables Median_Age (double), Migration_rate (double) and Youth_unemployment_rate (double) are measured variables in our case. Typically, the fixed variables are put in the beginning of the data set, while measured are put after them. So we will perform this small change to have our data perfectly tidy.

Also, since in the following tasks the region will be pointed to as "continent", we decided to rename the name of this variable. 

```{r}
df_vars <- df_vars %>%
  select(ISO, Country, region, sub.region, Income_class, Median_Age, Youth_unemployment_rate, Migration_rate)

names(df_vars)[names(df_vars) == "region"] <- "continent"

head(df_vars)
```
# g 
Make a frequency table for the status variable in the merged data set. Briefly comment on the results.

```{r}
income_status_frequency <- df_vars %>%
  count(Income_class, name = "frequency") %>%
  mutate(
    percentage = 100 * (frequency / sum(frequency))
  )

print(income_status_frequency)

```
From the generated table we can conclude that the high income countries are represented at most, while low income countries at least. The representation of lower-middle and upper-middle income countries are on the same level. Also, a lot of observations (about $19 \%$) are missing the value for the variable Income_class in our data set.

# h
What is the distribution of income status in the different continents? Compute the absolute frequencies as well as the relative frequency of status within each continent. Briefly comment on the results.

```{r}
income_distribution_within_continent <- df_vars %>%
  group_by(continent, Income_class) %>%
  summarise(frequency = n(), .groups = 'drop')

income_distribution_within_continent <- income_distribution_within_continent %>%
  group_by(continent) %>%
  mutate(relative_frequency = frequency / sum(frequency))

print(income_distribution_within_continent)


```
We can conclude that the low and lower-income classes are prevailing in Africa, while the biggest group in Europe and Oceania is high income class. In Asia the lower-middle income class is the biggest one followed by the upper-middle income class. In Americas the prevailing income class is upper-middle and in our data set there is no low income class countries represented. Also, it should be noted that each continent has some countries with no income class specification available (Africa $\approx 4\%$, Americas $\approx 16\%$, Asia $\approx 18\%$, Europe $\approx 13\%$ and Oceania $10\%$). In addition, some countries have values for variable Income_class, while no value for variable continent.  

# i 

From h. identify the countries which are the only ones in their respective group. Explain in few words the output.

```{r}
unique_countries_in__their_group <- df_vars %>%
  group_by(continent, Income_class) %>%
  filter(n() == 1) %>%
  ungroup()

print(unique_countries_in__their_group)
```

We can see that Ukraine is the only lower-middle income country in Europe, while the only high income country in Africa is Seychelles and the only low income country in Asia is Afghanistan. 

# j
For each continent count the number of sub-regions in the data set. How granular are the subcontinents that you employ in the analysis?

```{r}
sub_region_counts_for_continents <- df_vars %>%
  group_by(continent) %>%
  summarise(number_of_sub_regions = n_distinct(sub.region), .groups = 'drop')

print(sub_region_counts_for_continents)
```
It is seen that the Asia has the highest granularity, followed by Europe and Oceania, while Africa and Americas are represented only by two sub-regions in our data set. 


# k
Look at the frequency distribution of income status in the subregions of North- and South-Americas. Comment on the results.

```{r}
income_distribution_americas <- df_vars %>%
  filter(continent %in% "Americas") %>%
  group_by(sub.region, Income_class) %>%
  summarise(frequency = n(), .groups = 'drop') %>%
  group_by(sub.region) %>%
  mutate(relative_frequency = frequency / sum(frequency))

print(income_distribution_americas)

```
We obtained that in the Northern America all countries are high income and also we have one country with no specified income status. In the Southern America prevailing number of countries are upper-middle class income countries, around $24\%$ of countries are high income. 

# l.
Dig deeper into the low-middle income countries of the Americas. Which ones are they? Are they primarily small island states in the Caribbean? Comment.

```{r}
low_middle_income_americas <- df_vars %>%
  filter(continent %in% "Americas" & 
         Income_class == "LM")

print(low_middle_income_americas)

```
No, these countries are not island states (except Haiti), however they all except Bolivia have the coastal line. Also, they all are quite different in terms of the size. 

# m 
Create a table of average values for median age, youth unemployment rate and net migration rate separated into income status. Make sure that in the output, the ordering of the income classes is proper (i.e., L, LM, UM, H or the other way around). Briefly comment the results.

We will calculate it by ignoring the NA values. Also, we will exclude from the final data frame the values for the observations with no income class specified

```{r}
table_of_average_values <- df_vars %>%
  filter(!is.na(Income_class)) %>%
  mutate(Income_class = factor(Income_class, levels = c("L", "LM", "UM", "H"))) %>%  
  group_by(Income_class) %>%
  summarise(
    Avg_Median_Age = mean(Median_Age, na.rm = TRUE),
    Avg_Youth_Unemployment_Rate = mean(Youth_unemployment_rate, na.rm = TRUE),
    Avg_Migration_Rate = mean(Migration_rate, na.rm = TRUE),
    .groups = 'drop'
  )

print(table_of_average_values)

```

We can trace the growth of the average median age of the population with the drop of the status of country`s income. In addition, we observe that the highest average unemployment youth rate is experienced by the upper-middle status income countries. The high and lower-middle income countries have almost the same average unemployment youth rate and the indicator for low income status countries is the smallest. Average migration rate in countries of all statuses except high income status is negative, being the most negative in countries with upper-middle income status. 

# n
Look also at the standard deviation instead of the mean in m. Do you gain additional insights? Briefly comment the results.

```{r}
table_of_sd_values <- df_vars %>%
  filter(!is.na(Income_class)) %>%
  mutate(Income_class = factor(Income_class, levels = c("L", "LM", "UM", "H"))) %>%  
  group_by(Income_class) %>%
  summarise(
    Sd_Median_Age = sd(Median_Age, na.rm = TRUE),
    Sd_Youth_Unemployment_Rate = sd(Youth_unemployment_rate, na.rm = TRUE),
    Sd_Migration_Rate = sd(Migration_rate, na.rm = TRUE),
    .groups = 'drop'
  )

print(table_of_sd_values)
```

The standard deviation results do provide some additional insights. For the median age, we can see that the sd are relatively low across all income classes, meaning less variability, therefore similar age distributions.
The sd of youth unemployment rate is higher compared to the median age. This indicates a bigger variability in the rates, suggesting that employment conditions vary more among countries.
For the migration rate, the sd is lower for L, LM and H, and is the highest in UM class. This might show how the countries with UM income have more diverse migration patters. 


# o
Repeat the analysis in m. for each income status and continent combination. Discuss the results.

```{r}
table_of_average_values_continent <- df_vars %>%
  filter(!is.na(Income_class)) %>%
  mutate(Income_class = factor(Income_class, levels = c("L", "LM", "UM", "H"))) %>%  
  group_by(Income_class, continent) %>%
  summarise(
    Avg_Median_Age = mean(Median_Age, na.rm = TRUE),
    Avg_Youth_Unemployment_Rate = mean(Youth_unemployment_rate, na.rm = TRUE),
    Avg_Migration_Rate = mean(Migration_rate, na.rm = TRUE),
    .groups = 'drop'
  )

print(table_of_average_values_continent)

```
From the results we can observe the socioeconomic differences across the continents. We see that lower income countries have lower median ages, while higher income ones have older populations. Also, unemployment rates vary widely across regions, some having significantly higher rates than others. For L class, we have similar rates; for LM and UM, Africa has the highest rate; for H, Oceania has the highest rate. 
Furthermore, migration patters also differ. We have regions experimenting negative migration rates, while others having positive ones. L class has only negative rates; for LM, Europe has positive rate;for UM Africa has the positive one, while Oceania followed by Asia have the lowest rates; for H, we see only Oceania with negative rate.
Lastly, we have for each L, UM, H classes, an observation with NA values for the continent variable. This is is important to note as it might affect the accuracy of the results.  

# p
Identify countries which are doing well in terms of both youth unemployment and net migration rate (in the top 25% of their respective continent in terms of net migration rate and in the bottom 25% of their respective continent in terms of youth unemployment).

```{r}
top_countries <- df_vars %>%
  group_by(continent) %>%
  mutate(
    top_migration = quantile(Migration_rate, 0.75, na.rm = TRUE),
    bottom_unempoyment = quantile(Youth_unemployment_rate, 0.25, na.rm = TRUE)
  ) %>%
  filter(Migration_rate >= top_migration & Youth_unemployment_rate <= bottom_unempoyment) %>%
  select(Country, continent, Migration_rate, Youth_unemployment_rate)


top_countries
```
The results show diverse regions. Countries like Qatar, Switzerland, United Arab Emirates etc. present favorable results for both rates. Meanwhile, Africa countries, which have lower unemployment rates, have also lower migration rates.

# r
Export the final data set to a csv with “;” separator and “.” as a symbol for missing values; no rownames
should be included in the csv. Upload the .csv to TUWEL together with your .Rmd and .html (or .pdf).

```{r}

write.table(df_vars, file = "final_data.csv", sep = ";", na = ".", row.names = FALSE, col.names = TRUE)

```

